---
title: 'Exploratory Data Analysis: Course Project 2'
author: "Chaithanya Amin"
date: "13/09/2022"
output: html_document
---

### Fine particulate matter (PM2.5) is an ambient air pollutant for which there is strong evidence that it is harmful to human health. In the United States, the Environmental Protection Agency (EPA) is tasked with setting national ambient air quality standards for fine PM and for tracking the emissions of this pollutant into the atmosphere. Approximatly every 3 years, the EPA releases its database on emissions of PM2.5. This database is known as the National Emissions Inventory (NEI). 

### You can read more information about the NEI at the EPA National Emissions Inventory web site (http://www.epa.gov/ttn/chief/eiinformation.html).


## Data
### The data for this assignment are available from the course web site as a single zip file:
### https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip

### The zip file contains two files:

### PM2.5 Emissions Data (\color{red}{\verb|summarySCC_PM25.rds|}summarySCC_PM25.rds): This file contains a data frame with all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year.

## Instruction
### For each plot you should:
### Construct the plot and save it to a PNG file.

### Create a separate R code file (plot1.R, plot2.R, etc.) that constructs the corresponding plot, i.e. code in plot1.R constructs the plot1.png plot. Your code file should include code for reading the data so that the plot can be fully reproduced. You must also include the code that creates the PNG file. Only include the code for a single plot (i.e. plot1.R should only include code for producing plot1.png)

Upload the PNG file on the Assignment submission page

Copy and paste the R code from the corresponding R file into the text box at the appropriate point in the peer assessment.


## Install (if not yet) and load packages necessary for the analysis.
```{r}
install.packages("dplyr")
library(dplyr)
install.packages("ggplot2")
library(ggplot2)
```


## Downloading the data and saving it in the working directory
```{r}
filename <- "Data_for_Peer_Assessment.zip"
if (!file.exists(filename)) {
  download_url <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
  download.file(download_url, destfile = filename)
  unzip (zipfile = filename)
}
if (!exists("NEI")) {
  # print("Loading NEI Data, please wait.")
  NEI <- readRDS("summarySCC_PM25.rds") 
}
if (!exists("SCC")) {
  # print("Loading SCC Data.")
  SCC <- readRDS("Source_Classification_Code.rds")
}
```


## Assignment Questions and Answers.

### Question 1:
#### Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

#### Code performed:
```{r}
em_total <- aggregate(Emissions ~ year, NEI, FUN = sum)
color_range <- 2:(length(em_total$year)+1)
with(em_total, 
     barplot(height=Emissions/1000, names.arg = year, col = color_range, 
             xlab = "Year", ylab = expression('PM'[2.5]*' (Kilotons)'),
             main = expression('Total Emissions from PM'[2.5]*' (1999 to 2008)')))
```


### Question 2:
#### Have total emissions from PM2.5 decreased in the Baltimore City, Maryland ( fips == “24510”) from 1999 to 2008? Use the base plotting system to make a plot answering this question.

#### Code performed:
```{r}
bal <- NEI %>%
  filter(fips == "24510") %>%
  group_by(year) %>%
  summarise(Emissions = sum(Emissions))
with(bal,
       barplot(height=Emissions/1000, names.arg = year, col = color_range, 
               xlab = "Year", ylab = expression('PM'[2.5]*' (Kilotons)'),
               main = expression('Baltimore City, Maryland - Emissions (1999 to 2008)'))
)
```


### Question 3:
#### Of the four types of sources indicated by the (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008? Use the ggplot2 plotting system to make a plot answer this question.

#### Code performed:
```{r}
bal_em <- NEI %>%
  filter(fips == "24510") %>%
  group_by(year, type) %>%
  summarise(Emissions=sum(Emissions))
ggplot(data = bal_em, aes(x = factor(year), y = Emissions, fill = type, colore = "black")) +
  geom_bar(stat = "identity") + facet_grid(. ~ type) + 
  xlab("Year") + ylab(expression('PM'[2.5]*' Emission (kilotons')) +
  ggtitle("Baltimore City Emissions from Various Source Type") 
```


### Question 4:
#### Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?

#### Code performed:
```{r}
scc_coal <- SCC[grep("[Cc][Oo][Aa][Ll]", SCC$EI.Sector), "SCC"]
nei_coal <- NEI %>% filter(SCC %in% scc_coal)
coal_em_sum <- nei_coal %>% group_by(year) %>% summarise(Emissions = sum(Emissions))
ggplot(coal_em_sum, aes(x=year, y=round(Emissions/1000,2), label=round(Emissions/1000,2), fill=year)) +
  geom_bar(stat="identity") + ylab(expression('PM'[2.5]*' Emissions (Kilotons)')) + xlab("Year") +
  geom_label(aes(fill = year),colour = "white", fontface = "bold") +
  ggtitle("Coal Combustion Emissions Across U.S. (1999 to 2008).")
```


### Question 5:
#### How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?

#### Code performed:
```{r}
scc_vehicles <- SCC[grep("[Vv]ehicle", SCC$EI.Sector), "SCC"]
vehicle_em <- NEI %>% 
  filter(SCC %in% scc_vehicles & fips == "24510") %>%
  group_by(year) %>%
  summarise(Emissions = sum(Emissions))
  
ggplot(coal_em_sum, aes(x=year, y=round(Emissions/1000,2), label=round(Emissions/1000,2), fill=year)) +
  geom_bar(stat="identity") + ylab(expression('PM'[2.5]*' Emissions (Kilotons)')) + xlab("Year") +
  geom_label(aes(fill = year),colour = "white", fontface = "bold") +
  ggtitle("Baltimore City Vehicle Emissions (1999 to 2008).")
```


### Question 6:
#### Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == 06037). Which city has seen greater changes over time in motor vehicle emissions?

#### Code performed:
```{r}
baltcity_emissions<-summarise(group_by(filter(NEI, fips == "24510"& type == 'ON-ROAD'), year), Emissions=sum(Emissions))
l.a_emissions<-summarise(group_by(filter(NEI, fips == "06037"& type == 'ON-ROAD'), year), Emissions=sum(Emissions))
baltcity_emissions$County <- "Baltimore City, MD"
l.a_emissions$County <- "Los Angeles County, CA"
both.emissions <- rbind(baltcity_emissions, l.a_emissions)
ggplot(both.emissions, aes(x=factor(year), y=Emissions, fill=County,label = round(Emissions,2))) +
    geom_bar(stat="identity") + 
    facet_grid(County~., scales="free") +
    ylab(expression("total PM"[2.5]*" emissions in tons")) + 
    xlab("year") +
    ggtitle(expression("Motor vehicle emission variation in Baltimore and Los Angeles in tons"))+
    geom_label(aes(fill = County),colour = "white", fontface = "bold")
```
